---
title: "Week 7B: Experimenting with Matrix Models"
subtitle: 'EFB 370 - Spring 2022'
author: "Elie Gurarie and Nate Wehr"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r, message = FALSE, warning = FALSE}
require(expm)
require(gplots)
require(plyr)
require(magrittr)
require(utils)
library(DiagrammeR)
source("leslie_diagram.R")
library(magick)
```

In this lab, we are going to experiment with Matrix Models using the [Matrix Population Simulator 5000](https://egurarie.shinyapps.io/MatrixPopulationModels/).  You're basically going to mess around with it long enough to (try to) set up and run a bunch of matrix population processes.  For all the following processes, obtain (a) population growth $\lambda$ and (b) the  stationary distribution $N^*$.  And then answer remaining questions as they come up.  

# 1. Independent growth 

In lecture, we introduced the following very simple diagram: 

```{r}
par(mar = c(0,0,0,0))
M <- rbind(c(2,0),c(0,2))
leslie_diagram(M, stages = c("I", "II"))
```

Enter this diagram in to the simulator.  How does this process behave? What is $\lambda$?  What (if anything) is $N^{*}$?  How would you change this to have a stable population growth $\lambda = 1$? 

# 2. Criss-cross growth 

Here's a diagram where the life stages interact:

```{r}
par(mar = c(0,0,0,0))
M <- rbind(c(0,2),c(2,0))
leslie_diagram(M, stages = c("I", "II"))
```

How does this process behave? What is $\lambda$?  What (if anything) is $N^{*}$?  How would you change this to have a stable population growth $\lambda = 1$? 

# 3. Births and Deaths

The next example has a juvenile stage, all of whom survive to the adult stage, at which point adults produce 2 new juveniles. 

```{r}
par(mar = c(0,0,0,0))
M <- rbind(c(0,2),c(1,0))
leslie_diagram(M, stages = c("juv", "adult"))
```

What is $\lambda$?  What (if anything) is $N^{*}$?  What can you change to make the growth rate $\lambda = 1$?



# 4. A butterfly

Butterflies lay eggs that hatch into caterpillars and become adults.  Not all caterpillars become adults right away, leading to the following diagram:

```{r}
par(mar = c(0,0,0,0))
M <- rbind(c(0,0,60),c(0.1,0.2,0),c(0,.1,0))
leslie_diagram(M, stages = c("egg", "catepillar", "adult"))
```

What is $\lambda$?  What (if anything) is $N^{*}$?  

How could you tweak the **number of eggs to lay** to make the growth rate $\lambda = 1$?


# 5. Tasmanian Devil

Pre infectious facial tumor disease, the life history populaiton matrix for Tasmanian devils looked something like this:

```{r}
par(mar = c(0,0,0,0))
M.pre <- rbind(c(0,.1,.5,.7),
               c(.7,0,0,0),
               c(0,.8,0,0),
               c(0,0,.8,.5))
leslie_diagram(M.pre, stages = c("0", "1", "2", ">3"))

```

Answer the questions as before.  


## POST-DISEASE .... 

... the mortality of 2 year olds and 3 year olds became 90%.  This had an immediate impact on the population growth rate.  One way the population responsed was by increasing "precocial" births, i.e. 1 yr olds (and 2 year olds) that reproduced.  How high must that birth rate become to maintain the Tasmanian devil population? 