---
title: "Week 6: Fitting Logistic Growth Curves to Data"
subtitle: 'EFB 370 - Spring 2022'
author: "Nate Wehr and Elie Gurarie"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warnings = FALSE, message = FALSE)
```
```{r, echo = FALSE}
require(elieslides)
```


```{r, echo = FALSE, eval = FALSE}
data(gause_1934_book_f22)
a <- gause_1934_book_f22
#require(elieslides)
#ggplot(a, aes(Day, Volume_Species1, col = Treatment)) + geom_point()
a <- subset(gause_1934_book_f22, Treatment == "Pa")
b <- a[,c("Species1","Day","Volume_Species2")] %>% 
    plyr::rename(c(Species1 = "Species", Volume_Species2 = "Nt")) %>% 
    mutate(Nt1 = c(NA,Nt[-n]),
           dN = Nt-Nt1,
           r = dN/Nt1)
par(mfrow = c(1,3))
plot(Nt~Day, data = b)
plot(dN~Nt1, data = b)
plot(r~Nt1, data = b)
write.csv(b, file = "paramecium.csv", row.names = FALSE)
```


# Loading and exploring Paramecium data

Load data:

```{r}
p <- read.csv("paramecium.csv")
```

Look at the columns in this data:

```{r}
head(p)
```

What do the following commands do?

```{r, eval = FALSE}
nrow(p)
ncol(p)
dim(p)
names(p)
```

These data contain growth rates of paramecium (unicellular ciliates) raised in a petri dish with limited resources. 

The columns are: `Species`, `Day` of experiment, `Nt` - quantity of paramecium now, `Nt1` - quantity 1 time step ago $\Delta N_{t-1}$, and $r$ - per-capita growth rate, i.e. $\Delta N_t/N_t$.  


# Fit Method I: By Hand

Open the script file `logisticfitscript.R`.  It contains just the following code:

```{r}
p <- read.csv("paramecium.csv")
N.logistic <- function(x, N0, K, r0) K/(1 + ((K - N0)/N0)*exp(-r0*x))

# try different values!
N0 <- 10
K <- 150
r0 <- 1.5
plot(Nt~Day, data = p)
curve(N.logistic(x, N0, K, r0), add = TRUE, col = 2, lwd = 2)
```

The initial values guessed for `N0`, `K`, and `r0` are not very good.  How close can you get to a nice fit of the curve?  What is yuour best "by hand" guess of the parameters?  


# Fit method II: Using $r(N)$

## Ploting N, dN and r

Lets plot the three figures we talked about in class:  $N$ as a function of time, $\Delta N$ as a function of $N$, and $r$ as a function of $N$

```{r, echo = FALSE, par = -1, fig.height = 3, fig.width = 9}
pars(); par(mfrow = c(1,3))
plot(Nt~Day, data = p)
plot(dN~Nt1, data = p)
plot(r~Nt1, data = p)
```

These roughly show the patterns we expect of logistic regression.  

## Linear model of R

One form of the logistic growth model simply says that: 

$$ r(N) = r_0 - {r_0 \over K} N$$

So if we do a straightforward linear regression, we can estimate both $r$ and $K$:

```{r}
lm(r ~ Nt1, data = p)
```

This suggests that r0 is roughly 1.09.  We write:  $$\widehat{r_0} = 1.09$$ 

What does the slope tell us?  Well - we can convert that to a carrying capacity, since:  $$\beta = -r_0/K.$$

Solving for $\widehat{K}$ we get: $$\widehat{K} = 1.088 / 0.0052 = 209.1$$

## Check the fit

Is this estimate any good? Try putting these numbers into the fitting code:

```{r, eval = FALSE}
N0 <- 10
K <- 209
r0 <- 1.088
plot(Nt~Day, data = p)
curve(N.logistic(x, N0, K, r0), add = TRUE, col = 2, lwd = 2)
```

We need to adjust $N_0$ - but otherwise not too bad. 

## Standard errors

Note - we can - again - get standard errors around these coefficients: 

```{r}
r.lm <- lm(r ~ Day, data = p)
summary(r.lm)
```

This is pretty good!  Especially for the $r_0$ variable - basically $\widehat{r} 1.2 \pm 0.42$. But, two (small) problems: 

- hard to convert these to a standard error around $K$
- no way to obatin an estimate for $N_0$ from this model 


# Fit Method III:  Non-linear least squares

One method of fitting a curve is to use the "least squares" method, i.e. find the parameters for which the squared devation of the data from the curve is minimized.  This is an *optimization* problem that the computer solves numerically. 

The function that does this in R is called `nls()`.  It is "non-linear", unlike the regression model above.  To do this, we have to use our function for the solution of the logistic curve and provide initial values.  But the basic syntax is similar in that the response is `Nt` and the predictor is the logistic function of time (`Day`),  with the parameters to be estimated following that `Day`.  And the formula symbol is always `~`. 

```{r}
p <- read.csv("paramecium.csv")
N.logistic <- function(x, N0, K, r0) K/(1 + ((K - N0)/N0)*exp(-r0*x))

nls(Nt ~ N.logistic(Day, N0, K, r0), data = p, 
    start = list(N0 = 1, K = 200, r0 = 0.75))
```

As usual, we can save the results to an object and get a summary table with the fits:

```{r}
summary(nls(Nt ~ N.logistic(Day, N0, K, r0), data = p, 
    start = list(N0 = 1, K = 200, r0 = 0.75)))
```


<font color = "red">

# Additional homework problem 4

> Report the estimates of a fitted logistic curve to the Paramecium data, with confidence intervals and produce a plot of the fitted curve to the data.  

</font>