---
title: "Population Growth With Stochasticity"
author: '[Elie Gurarie](https://eligurarie.github.io/)'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
runtime: shiny
subtitle: 'EFB 370: Population Ecology'
resource_files:
- environmentalstochasticity.R
- demographicstochasticity.R
---
<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
#output: html_document
```


## I. Demographic Stochasticity

In this simulation tool, a given number of individuals ($N_0$) undergo a stochastic, discrete time, birth death process.  If death occurs before birth, each individual has a fixed probability of dying ($p_d$) and remaining survivors have a fixed probability of reproducing ($p_b$).  If birth occurs before death, the process is reversed.  

```{r}
source("demographicstochasticity.R")
shinyApp(ui=ui, server=server, options = list(height = 800))
```

## II. Environmental Stochasticity

Here, the discrete change in annual growth is given by the following function:

$$N_{t+1} = N_t \exp(R)$$
Where $R$ is a random variable distributed as:
$$R \sim {\cal N}(\mu_r, \sigma_r)$$

i.e., the average growth rate is $\mu_r$, but with interannual variation.  This is *almost* the same as: 
$$N(t) = N_0 \exp(R\,t)$$


According to theory, if $\sigma_r^2 > 2 \mu_r$, the population will "almost certainly" go to extinction.

```{r}
source("environmentalstochasticity.R")
shinyApp(ui=ui, server=server, options = list(height = 800))
```

